---
title: "Dengue Watch"
execute: 
  cache: refresh
---

```{r}
#| label: libraries

library(knitr)
library(crosstalk)

library(ggiraph)
library(ggplot2)
library(dygraphs)
library(xts)
library(plotly)
library(patchwork)

library(data.table)
options(datatable.week = "sequential")
```

```{r}
#| label: ggiraph options


set_girafe_defaults(
  opts_sizing = opts_sizing(rescale = TRUE, width = 1),
  opts_tooltip = opts_tooltip(
    opacity = 1, offx = 20, offy = -10,
    use_fill = TRUE, use_stroke = FALSE,
    delay_mouseout = 1000
  ),
  opts_toolbar = opts_toolbar(
    saveaspng = FALSE,
    hidden = c("selection", "zoom", "misc")
  )
)
```

```{r}
#| label: load data


denguDf <- fread("data/dengue_data.csv")
distMap <- sf::st_read("data/lka_ADM2_simplified.geojson", quiet = TRUE)
distPop <- fread("data/district_population.csv")
```

```{r}
#| label: claan data


names(distMap)[[1]] <- "district"
distMap <- distMap[c("district", "geometry")]
distMap$district <- with(distMap, sub(" District", "", district))

## ---- match district names to map
denguDf[, district := fcase(
  district == "Kalmune", "Ampara",
  district == "Hambanthota", "Hambantota",
  district == "NuwaraEliya", "Nuwara Eliya",
  default = district
)][, cases := sum(cases), by = .(district, start.date, end.date)]

## ---- create additional columns
denguDf[, let(
  start.date  = as.IDate(start.date),
  end.date    = as.IDate(end.date),
  year        = year(start.date),
  month       = format(start.date, "%B"),
  daymon      = mday(start.date),
  epiweek     = week(end.date)
)]

month_abb <-  c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                              "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
pop_lookup <- setNames(distPop$pop_000, distPop$district)

sharedDengue <- SharedData$new(denguDf)
```

```{r}
#| label: color pallets

pal <- list( 
  c(
    "#f7f4f9", "#e7e1ef", "#D4B9DA", "#c994c7", "#df65b0",
    "#e7298a", "#ce1256", "#980043", "#67001f"
  )
)
```

# Overview of 2026 {orientation="columns"}

## Map {.tabset width="35%"}

```{r}
#| title: Dengue Incident by District - 2026
#| fig-width: !expr 12 * 0.56
#| fig-height: 12
#| padding: 0px

local({
  df <- denguDf[year == 2026, .(cases = sum(cases)), by = district]
  df[, incident := cases / pop_lookup[district] * 100]

  map <- merge(distMap, df, by = "district")
  map$tooltip <- with(map, sprintf("%s:\ntotal cases: %i", district, cases))

    fig <- ggplot(map) +
      geom_sf_interactive(
        aes(fill = incident, data_id = incident, tooltip = tooltip, tootltip_color = "black"),
        lwd = 0.05, color = "black") +
      scale_fill_continuous(
        name = "",
        palette = pal[[1]],
        breaks = seq(30, 300, 30),
        guide = guide_legend(
          keyheight = unit(5, units = "pt"),
          keywidth = unit(10, units = "pt"),
          label.position = "bottom",
          title.position = "top",
          nrow = 1)) +
      labs(caption = "Incident Rates based on 2025 mid year population") +
      theme_void() +
      theme(
        aspect.ratio = 12.5 / 7,
        legend.position = c(0.65, 0.95),
        legend.text = element_text(size = 10),
        panel.background = element_blank(),
        plot.caption = element_text(size = 10),
        plot.caption.position = "panel",
        plot.margin = margin(0, 0, 0, 0, unit = "pt"))

    girafe(fig)
  
})
```

```{r}
#| title: Total Reported Cases - 2026
#| fig-width: !expr 12 * 0.56
#| fig-height: 12
#| padding: 0px

local({
  df <- denguDf[year == 202, .(cases = sum(cases)), by = district]

  map <- merge(distMap, df, by = "district")
  map$tooltip <- with(map, sprintf("%s:\ntotal cases: %i", district, cases))

  fig <- ggplot(map) +
    geom_sf_interactive(
      aes(fill = cases, data_id = cases / 1000, tooltip = tooltip),
      lwd = 0.05, color = "black") +
    scale_fill_continuous(
      name = "",
      palette = pal[[1]],
      guide = guide_legend(
        keyheight = unit(5, units = "pt"),
        keywidth = unit(10, units = "pt"),
        label.position = "bottom",
        title.position = "top",
        nrow = 1)) +
    labs(caption = "Incident Rates based on 2025 mid year population") +
    theme_void() +
    theme(
      aspect.ratio = 12.5 / 7,
      legend.position = c(0.65, 0.95),
      legend.text = element_text(size = 10),
      panel.background = element_blank(),
      plot.caption = element_text(size = 10),
      plot.caption.position = "panel",
      plot.margin = margin(0, 0, 0, 0, unit = "pt"))

    girafe(fig)
  
})
```

## Summary {width="70%"}

### Value Boxes {height="20%"}

```{r}
#| content: valuebox

list(
  title = "Total Reported cases",
  value = paste0(denguDf[year == 2026, .(sum(cases))], "+"),
  
  icon = "virus", color = "warning"
)
```

```{r}
#| content: valuebox
#| title: "Total Sales"
#| icon: "cart"
#| color: "success"
list(
  value = 1245
)
```

```{r}
#| content: valuebox
#| title: "Total Sales"
#| icon: "cart"
#| color: "success"
list(
  value = 1245
)
```

### Details {height="80%"}

#### 2026 in Summary {width="60%"}

::: {.card-body height="30%"}
Sri Lanka has recorded over **56,000** dengue cases in 2026, a marked increase over 2025, driven largely
by **DENV-2**. The Western Province (**Colombo**, **Gampaha**, **Kalutara**) remains the hardest-hit region due to
population density and mosquito breeding sites.
:::

##### Summary plots {height="70%"}

###### {width="50%"}

```{r}
local({
  df <- denguDf[year == 2026, .(cases = sum(cases, na.rm = TRUE)), by = month]
  df[, tooltip := sprintf("Reported cases: %i", cases)]
  
  fig <- ggplot(df) +
      geom_col_interactive(
        aes(x = factor(month, month.name), y = cases, data_id = cases, tooltip = tooltip),
        fill = "pink", color = "black", lwd = 0.3) +
      geom_hline(yintercept = -2, lwd = 1) +
      geom_text(aes(month, cases + cases/100 * 5, label = month))+
      labs(title = "Monthly Reported Dengue Cases - 2026") +
      theme_void() +
      theme(
        plot.title.position = "panel", 
        text = element_text(size = 16)
      )
  
  girafe(fig)
})
```

###### {width="50%"}

```{r}
#| out-width: "100%"


local({
  df <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(district, year, month)]
  df[, month_num := match(month, month.name)]
  df[, date := as.Date(sprintf("%d-%02d-01", year, month_num))]

  ## fix a stacking order so districts don't swap layers over time
  order_lvls <- df[, .(total = sum(cases)), by = district][order(-total), district]
  df[, district := factor(district, levels = order_lvls)]
  setorder(df, date, district)

  ## silhouette (centered) streamgraph baseline
  df[, cum_cases   := cumsum(cases), by = date]
  df[, total_cases := sum(cases), by = date]
  df[, let(
    ymax = cum_cases - total_cases / 2,
    ymin = cum_cases - cases - total_cases / 2
  )]
  df[, tooltip := sprintf("%s\n%s %d: %i cases", district, month, year, cases)]

  girafe({
    ggplot(df, aes(date, group = district)) +
      geom_ribbon_interactive(
        aes(ymin = ymin, ymax = ymax, fill = district, data_id = district, tooltip = tooltip),
        color = "white", lwd = 0.1
      ) +
      scale_fill_viridis_d(option = "turbo") +
      labs(title = "Streamgraph of Dengue Cases by District", x = "", y = "", fill = "District") +
      theme_minimal() +
      theme(axis.text.y = element_blank(), panel.grid = element_blank())
  })
})
```


#### Heat map {width="40%"}

```{r}
local({
  df <- denguDf[year == 2026, .(cases = sum(cases, na.rm = TRUE)), by = .(district, month)]
  df[, tooltip := sprintf("Reported Cases: %i", cases)]
  
  fig <- ggplot(df) +
    geom_tile_interactive(
      aes(factor(month, month.name), y = reorder(district, cases, sum),
          fill = cases, data_id = cases, tooltip = tooltip), color = "white") +
    scale_fill_continuous(
      palette = pal[[1]],
      guide = guide_colorbar(title = "Cases" )) +
    labs(title = "District x Month Heatmap of Dengue Cases (2026)",
         x = "", y = "", fill = "Cases")
  
  girafe(fig)
})
```

# Changes Over Time {.columns}

## Time Series {.tabset}

###

```{r}
local({
  df <- denguDf[year == 2026, .(cases = sum(cases, na.rm = TRUE)), by = .(district, month)]

  ## add a "Total" pseudo-month so it's selectable alongside individual months
  total <- df[, .(cases = sum(cases)), by = district]
  total[, month := "Total"]
  df_all <- rbindlist(list(df, total), use.names = TRUE)
  df_all[, month := factor(month, levels = c("Total", month.name))]

  ## fix district order (by total cases) so bars don't reshuffle per selection
  district_order <- df_all[month == "Total"][order(-cases), district]
  df_all[, district := factor(district, levels = district_order)]

  sd <- SharedData$new(df_all, key = ~month, group = "month_filter")

  p <- ggplot(sd, aes(district, cases,
                      text = sprintf("%s\n%s: %i cases", district, month, cases))) +
    geom_col_interactive(fill = "darkred", aes(data_id = cases)) +
    labs(title = "Dengue Cases by District (2026)", x = "", y = "Cases") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  bscols(
    widths = c(3, 9),
    filter_select("month_sel", "Month", sd, ~month, multiple = FALSE),
    girafe(p)
  )
})
```


### Monthly Dengue Cases - Since 2006 {.rows}

:::{.card-body}
The monthly dengue cases show clear changes over time, with both seasonal patterns and occasional 
large outbreaks. From **2007 to 2015**, the number of reported cases remained relatively low, with 
only a few moderate increases. However, cases began to rise after **2016**, reaching a major peak 
in **2017**, which recorded the highest monthly number of dengue cases.

Although the number of cases decreased after the 2017 outbreak, dengue continued to occur every year
with noticeable increases, particularly in **2019**, **2023**, and **2024**.
Overall, the results show that dengue remains a recurring public health concern in Sri Lanka, 
with the number of cases varying from year to year and increasing during certain seasons.
:::

```{r}
#| out-width: "100%"

local({
  df <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(year, month)]
  df[, let(month_num = match(month, month.name))]
  df[, date := as.Date(sprintf("%i-%02i-01", year, month_num))]
  df[, tooltip := sprintf("%s %i: %i cases", month, year, cases)]

  
  xts(df$cases, order.by = df$date) |>
    dygraph() |> 
    dyOptions(
      drawPoints = TRUE, pointSize = 4,
      strokeWidth = 2,
      drawGrid = FALSE, labelsKMB = TRUE)
})
```

### Seasonal Decomposition
```{r}
local({
  df <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(year, month)]
  df[, month_num := match(month, month.name)]
  setorder(df, year, month_num)

  ts_cases <- ts(df$cases, start = c(df$year[1], df$month_num[1]), frequency = 12)
  decomp <- decompose(ts_cases, type = "additive")

  decomp_dt <- data.table(
    date     = as.Date(sprintf("%i-%02i-01", df$year, df$month_num)),
    observed = as.numeric(decomp$x),
    trend    = as.numeric(decomp$trend),
    seasonal = as.numeric(decomp$seasonal),
    random   = as.numeric(decomp$random)
  )
  decomp_long <- melt(decomp_dt, id.vars = "date", variable.name = "component", value.name = "value")
  decomp_long[, component := factor(component, levels = c("observed", "trend", "seasonal", "random"))]
  decomp_long$tooltip <- with(decomp_long, sprintf("%s: %.1f", component, value))

  fig1 <- ggplot(decomp_long, aes(date, value)) +
      geom_line_interactive(aes(data_id = component, tooltip = tooltip), color = "darkred") +
      facet_wrap(~component, scales = "free_y", ncol = 1) +
      labs(title = "Decomposition of Monthly Dengue Cases", x = "", y = "") +
      theme_minimal() 
  girafe({
   fig1
  })
})
```

### {.rows title="Seasonal Index by Month"}

#### {height="20%"}
::: {.card-body}
The seasonal index demonstrates a distinct monthly fluctuation in dengue cases. A seasonal index 
exceeding **1.0** signifies months with higher-than-average dengue occurrences, while values falling 
below **1.0** indicate lower-than-average incidence. Typically, dengue cases are above average 
in **January, June, July, November, and December**, with **July** showing the highest seasonal 
index at around **1.65**. In contrast, **February through May** and **August through October** 
reflect below-average dengue activity, with **September** registering the lowest seasonal index of 
approximately **0.65**.
:::

#### {height="80%"}

##### {.tabset width="70%"}

```{r}
#| title: Seasonal Index by Month


local({
  idx <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(year, month)][
    , .(avg_cases = mean(cases, na.rm = TRUE)), by = month
  ]
  idx[, seasonal_index := avg_cases / mean(avg_cases)][order(month)]
  idx[, tooltip := sprintf("%s:\nindex %.2f", month, seasonal_index)]

  girafe({
    ggplot(idx, aes(factor(month, month.name), seasonal_index)) +
      geom_col_interactive(
        aes(data_id = seasonal_index, tooltip = tooltip),
        fill = "darkred", color = "black", lwd = 0.3, width = 0.7
      ) +
      geom_hline(yintercept = 1, linetype = "dashed") +
      geom_hline(yintercept = 0, linewidth = 1) +
      guides(x = guide_axis(n.dodge = 2)) +
      labs(title = "", x = "", y = "Index") +
      theme(
        aspect.ratio = 0.5,
        plot.background = element_blank()
      )
  })
})
```

```{r}
#| title: Data


local({
  idx <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(year, month)]
  idx <- idx[, .(avg_cases = mean(cases, na.rm = TRUE)), by = month]
  idx[, seasonal_index := avg_cases / mean(avg_cases)]
  setorder(idx, month)

  kable(idx[, .(
      Month = month, 
      `Average Cases` = round(avg_cases, 1), 
      `Seasonal Index` = round(seasonal_index, 2)
      )])
})

```

##### {width="30%"}
::: {.card-body}
The identified seasonal trend reveals that _dengue transmission is mainly concentrated during
certain times of the year_, rather than being evenly distributed throughout all months. 
The significant peak observed in **July**, along with the heightened seasonal indices in **June**, **November**, 
and **December**, may correlate with the seasonal rainfall patterns in Sri Lanka.

Increased rainfall can foster suitable breeding environments for mosquitoes by providing more stagnant water, which 
in turn boosts vector populations and the likelihood of dengue transmission. In contrast, the lower 
seasonal indices noted during March to May and September to October may indicate periods with less 
conducive environmental conditions for mosquito breeding.
:::

###

```{r}
#| title: Year over Year

local({
  df <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(year, month)]
  df[, month_num := match(month, month.name)]
  df$tooltip <- with(df, sprintf("%s %i:\n%i cases", month, year, cases))

  girafe({
    ggplot(df, aes(factor(month, month.name), cases, group = year, color = factor(year))) +
      geom_line() +
      geom_point_interactive(aes(data_id = cases, tooltip = tooltip), size = 1.5) +
      labs(title = "Year-over-Year Monthly Dengue Cases", x = "", y = "Cases", color = "Year") +
      theme_minimal()
  })
})
```

```{r}
#| title: Autocorrelation

local({
  df <- denguDf[, .(cases = sum(cases, na.rm = TRUE)), by = .(year, month)]
  df[, month_num := match(month, month.name)]
  setorder(df, year, month_num)

  ts_cases <- ts(df$cases, start = c(df$year[1], df$month_num[1]), frequency = 12)
  acf_vals <- acf(ts_cases, plot = FALSE, lag.max = 24)
  acf_dt <- data.table(lag = acf_vals$lag[, 1, 1], acf = acf_vals$acf[, 1, 1])
  ci <- qnorm((1 + 0.95) / 2) / sqrt(acf_vals$n.used)
  acf_dt$tooltip <- with(acf_dt, sprintf("lag %f:\nacf %.2f", lag, acf))

  girafe({
    ggplot(acf_dt, aes(lag, acf)) +
      geom_col_interactive(aes(data_id = acf, tooltip = tooltip), fill = "darkred", width = 0.3) +
      geom_hline(yintercept = c(-ci, ci), linetype = "dashed", color = "steelblue") +
      labs(title = "Autocorrelation of Monthly Dengue Cases", x = "Lag (months)", y = "ACF") +
      theme_minimal()
  })
})
```
