---
title: "Dengue Watch"
---


```{r}
#| label: Plot Script

library(ggiraph)

source("rscripts/analysis.R")

set_girafe_defaults(
  opts_sizing = opts_sizing(rescale = TRUE, width = 1),
  opts_toolbar = opts_toolbar(
    saveaspng = FALSE,
    hidden = c("selection", "zoom", "misc")
  )
)
```

# Overview of 2026 {orientation="columns"}

## Column {.tabset width="40%"}

```{r}
#| title: Incident Rates - 2026
#| label: Map of Incident Rates
#| fig-width: !expr 12 * 0.56
#| fig-height: 12
#| padding: 0px

girafe(fig_1)
``` 

```{r}
#| title: Monthly & District Breakdown - 2026

fig_12

fig_incidence_rank
```

## Column {width="60%"}

### Row {height="5%"}

```{r}
#| label: valuebox-1
#| content: valuebox

list(
  title = "Total Reported Cases",
  value = "56000+",
  color = "warning", icon = "virus"
)
```
```{r}
#| label: valuebox-2
#| content: valuebox

list(
  value = 1245
)
```

### Row {height="20%}

#### {.columns}

##### Column {width="70%"}

:::{.card}
Dengue transmission in 2026 shows clear geographic variation, 
with relatively higher incidence observed in several southern and south-western districts. 
During **January–May**, **Colombo**, **Galle** and **Gampaha** remained the districts
with the highest reported dengue incident rates, although changes in the rankings of 
other districts suggest shifts in dengue activity compared with the same period in 2025.
:::


### Row {.tabset height="70%"}

```{r}
#| title: Outbreak Detection

fig_outbreak
```

```{r}
fig_4
```

```{r}
fig_cumulative
```

```{r}

fig_boxplot
```

```{r}
fig_growth_rate
```

# Time Series Analysis {orientation="columns"}

## Column {.tabset}

### Tab Row 

#### {.columns}

##### Column

```{r}
#| out-width: "100%"

fig_55
```

#####

```{r}
#| out-width: "100%"

fig_5
```

:::{.card-body}
The seasonal index indicates that dengue cases follow a clear seasonal pattern
throughout the year rather than being evenly distributed. July has the highest
seasonal index, showing that dengue cases are substantially higher than the annual
monthly average during this period. Elevated indices in January, June, November,
and December also suggest increased transmission, while March, April, September,
and October consistently experience below-average case levels. Overall, the results
highlight a mid-year peak with a secondary increase towards the end of the year,
suggesting that dengue prevention and control measures should be strengthened
before these high-risk periods.
:::

### Row {.tabset}

```{r}
#| title: 8-Week Forecast

fig_forecast
```

```{r}
#| title: Weekly YoY Comparison

fig_3
```

```{r}
#| title: Autocorrelation

fig_6
```

# District Breakdown

## {.tabset}

```{r}
#| title: Top 10 Districts (2025 vs 2026)

fig_2
```

```{r}
fig_small_multiples
```

```{r}
fig_district_corr
```


# Data Sources

:::{.card title="Data Sources"}
Historical case data (pre-2026): Talagala T (2026). *denguedatahub: A Tidy Format
Dataset of Dengue by Country*. R package version 4.1.1.
[thiyangt.github.io/denguedatahubweb](https://thiyangt.github.io/denguedatahubweb/)

2026 Dengue data: Sri Lanka Epidemiology Unit, weekly epidemiological reports.

Population data: Department of Census and Statistics, Sri Lanka (2025). *Mid-year Population Estimates
by District*. #link("http://www.statistics.gov.lk/")
:::