---
title: "Dengue Watch"
---


```{r}
#| label: Plot Script

library(ggiraph)

source("rscripts/analysis.R")

set_girafe_defaults(
  opts_sizing = opts_sizing(rescale = TRUE, width = 1),
  opts_toolbar = opts_toolbar(
    saveaspng = FALSE,
    hidden = c("selection", "zoom", "misc")
  )
)
```

# Overview of 2026 {orientation="columns"}

## Column {.tabset width="40%"}

```{r}
#| title: Incident Rates - 2026
#| label: Map of Incident Rates
#| fig-width: !expr 12 * 0.56
#| fig-height: 12
#| padding: 0px

girafe(fig_1)
``` 

```{r}
#| title: Monthly & District Breakdown - 2026

fig_12

fig_13
```

## Column {width="60%"}

### Row {height="5%"}

```{r}
#| label: valuebox-1
#| content: valuebox

list(
  title = "Total Reported Cases",
  value = "56000+",
  color = "warning", icon = "virus"
)
```
```{r}
#| label: valuebox-2
#| content: valuebox

list(
  value = 1245
)
```

### Row {height="20%}

:::{.card}
Dengue transmission in 2026 shows clear geographic variation, 
with relatively higher incidence observed in several southern and south-western districts. 
During **January–May**, **Colombo**, **Galle** and **Gampaha** remained the districts
with the highest reported dengue incident rates, although changes in the rankings of 
other districts suggest shifts in dengue activity compared with the same period in 2025.
:::

### Row {height="70%"}

```{r}
fig_4
```

# Time Series Analysis {orientation="columns"}

## Column {.tabset}

### Tab Row 

#### {.columns}

##### Column
:::{.card-body}
```{r}
#| out-width: "100%"

fig_55
```
The decomposition reveals a clear seasonal cycle in dengue cases, a generally increasing
long-term trend, and several irregular outbreaks. While seasonal patterns remain
consistent from year to year, large epidemic peaks—particularly around 2017—are driven
by factors beyond normal seasonal variation, highlighting the importance of continuous
surveillance and timely public health interventions.
:::

#####

```{r}
#| out-width: "100%"

fig_5
```

:::{.card-body}
The seasonal index indicates that dengue cases follow a clear seasonal pattern
throughout the year rather than being evenly distributed. July has the highest
seasonal index, showing that dengue cases are substantially higher than the annual
monthly average during this period. Elevated indices in January, June, November,
and December also suggest increased transmission, while March, April, September,
and October consistently experience below-average case levels. Overall, the results
highlight a mid-year peak with a secondary increase towards the end of the year,
suggesting that dengue prevention and control measures should be strengthened
before these high-risk periods.
:::