---
title: "Dengue Watch"
---


```{r}
#| label: Plot Script

library(ggiraph)

source("rscripts/analysis.R")

set_girafe_defaults(
  opts_sizing = opts_sizing(rescale = TRUE, width = 1),
  opts_toolbar = opts_toolbar(
    saveaspng = FALSE,
    hidden = c("selection", "zoom", "misc")
  )
)
```

# Overview of 2026 {orientation="columns"}

## Column {width="40%"}

```{r, cache=TRUE}
#| label: Map of Incident Rates
#| fig-width: !expr 12 * 0.56
#| fig-height: 12
#| padding: 0px

girafe(fig_1)
```

## Column {width="60%"}

### Row {height="5%"}

```{r}
#| label: valuebox-1
#| content: valuebox

list(
  title = "Total Reported Cases",
  value = "56000+",
  color = "warning", icon = "virus"
)
```
```{r}
#| label: valuebox-2
#| content: valuebox

list(
  value = 1245
)
```

### Row {height="5%"}

```{r}
#| label: valuebox-3
#| content: valuebox

list(
  title = "Total Reported Cases",
  value = "56000+",
  color = "warning", icon = "virus"
)
```
```{r}
#| label: valuebox-4
#| content: valuebox

list(
  value = 1245
)
```
```{r}
#| label: valuebox-5
#| content: valuebox

list(
  value = 1245
)
```

### Row {height="20%}

:::{.card}
Dengue transmission in 2026 shows clear geographic variation, with relatively higher incidence observed in several southern and south-western districts. During **January–May**, **Colombo**, **Galle** and **Gampaha** remained the districts with the highest reported dengue incident rates, although changes in the rankings of other districts suggest shifts in dengue activity compared with the same period in 2025.

The difference between incident rates and case counts also highlights the importance of considering population size when interpreting district-level disease burden.
:::

### Row {height="70%"}

#### Column {width="50%"}

```{r, cache=TRUE}
#| label: Top 10 Districts
#| fig-height: 6

girafe(fig_2)
```

#### Column {width="50%"}

```{r}
plot(mtcars$cyl)
```